Metal Mixing and Ejection in Dwarf Galaxies is Dependent on Nucleosynthetic Source

### 2018-07-30 main.tex
rounding out Benoits comments. updating citations in introduction

\section{Introduction}
Understanding galactic chemical evolution for all metal species across galaxy mass scales remains one of the most challenging aspects of modeling galaxy evolution (\textit{references}). One of the most pressing difficulties is a lack of understanding in exactly how metals propagate from their injection sites (from stellar winds or supernovae) and mix through the phases of the interstellar medium (ISM) into star forming gas.
%
% get a sentence with a better punch at the end
%

One-zone chemical evolution models assume homoegeneous mixing of metals from recent star formation into gas available for future star formation (references). While this incorrectly assumes that metal mixing in the ISM plays no role in delaying future enrichment in star formation, how to properly account for this process is poorly understood. This is in large part because both these models and large-scale cosmological simulations lack the necessary fidelity to capture the detailed, multi-phase mixing process of metals in the ISM directly. Recent hydrodynamics simulations have employed parameteric models to account for unresolved sub-grid metal mixing (\textit{need references}), which plays an important role in determining the chemical properties of galaxies (\textit{need references}). Consequently, following metal mixing, if treated in these models, requires assumptions about how metals mix within the ISM and how metals couple to galactic outflows. When metals are tracked individually, as opposed to a global metallicity field, their injection, mixing, and outflow properties are often treated uniformly. However, these processes are critical components in understanding galactic chemical evolution. Detailed hydrodynamics simulations incorporating a realistic multi-phase ISM and detailed stellar feedback are required to understand metal mixing and metal outflows \citep[e.g.][]{Armillotta2018}. In addition, it remains to be seen to what degree, if at all, metals of different nucleosynthetic origins may couple differently to the ISM.

@@

It is becoming increasingly important to develop a concrete theoretical understanding of galactic chemical evolution as a result of multiple, recent observational campaigns to probe detailed stellar abundances in our Milky Way and the Local Group. (\textit{ references to APOGEE, GAIA, GALAH, RAVE, TGAS}). Stellar abundances are directly imprinted with the enrichment pattern of their star forming cloud, whose chemical properties are determined by the process of turbulent metal mixing in the ISM. The degree to which we can associate "chemically tagged" stars as co-eval depends directly upon our understanding of metal enrichment and metal mixing in the ISM. However, this understanding is critical for using these observations to further deduce properties of a galaxy's evolutionary history.

@@

%mm \textit
\paragraph{Stellar Feedback and Stellar Yields} We track the feedback and yields of 15 metal species for each star individually. Stars between 8 M$_{\odot} < M_{*} < 25 M_{\odot}$ explode as core collapse supernovae at the end of their life, injecting their mass and 10$^{51}$~erg of thermal energy into a spherical region with radius of 5.4~pc, or 3 times the maximum resolution. Stars above this mass are assumed to direct collapse with no mass or energy injection. For all stars above 8~M$_{\odot}$ we follow their stellar winds assuming continuous mass loss over their lifetimes, their LW and FUV radiation as optically thin radiation which contributes to H$_2$ dissociation and photoelectric heating respectively, and their H~{\sc i} and He~{\sc i} ionizing radiation using an adaptive ray-tracing radiative transfer method \citep{WiseAbel2011}. We interpolate over the OSTAR2002 \citep{Lanz2003} grid to set the luminosities of each of these stars. Low mass stars, $M_{*} < 8~M_{\odot}$, do not produce feedback during their main sequence lifetimes, but end their lives injecting a short-lived, low velocity (10 km~s$^{-1}$) AGB wind \citep{Goldman2017}. Stars with $3 < M_{*} < 8$ are tracked after their death as possible Type Ia supernova progenitors, using a delay time distribution model to assign when (if at all) they will explode as a Type Ia, injecting 10$^{51}$~erg of thermal energy with yields. Stellar yields are computed using the NuGrid stellar yield database \citep{Ritter2017} for all stars with $M_{*} < 25~M_{\odot}$, \textbf{Slemer et. al. in prep} for the stellar winds of stars with $M_{*} > 25~M_{\odot}$, and \cite{Thielemann1986} for Type Ia supernovae.

@@

The only physics that separates the dynamical evolution of these elements in our simulations is the individual sources of enrichment: AGB winds (stars less than 8~M$_{\odot}$), stellar winds (stars above 8~M$_{\odot}$), core collapse supernovae, and Type Ia supernovae. These channels differ in: 1) how long after a given star formation event they occur, 2), by consequence, the typical ISM properties in which they occur, and 3) how much energy is associated with each event, which determines ejecta temperature and velocity. To understand where each of the elements in Figure~\ref{fig:species_fractions} originated, we show the mass fraction of metals produced through each channel in Figure~\ref{fig:species_sources}. Core collapse supernovae are responsible for over 90\% of the total metal enrichment in our galaxy, but this is clearly not true for all elements. In particular, a majority of N (74\%) and Ba (65\%) are ejected by AGB winds; both are retained at a higher rate than the rest of the metals. That this behavior exists for N and Ba in our simulations is dependent upon the metallicity of our galaxy and choice of stellar yield tables, but we can generalize this result to say that, for any assumed set of yields, low mass galaxies should more easily retain \textit{any} elements synthesized predominately in AGB winds, as compared to elements synthesized through supernovae.\footnote{One might expect that much of the N and Ba that is ejected by galactic winds is comprised mostly of the $\sim$30\% of each species produced through the stellar winds of more massive stars and supernovae. However, this cannot be
%mm determined
    proven
as we lack Lagrangian information about individual gas elements.}
% AE: Actually if there are significant isotope variations between nucleosynthetic sources (i.e. N produced in AGB winds is a different isotope than the trace N produced in SNe) then that could be a very cool thing to distinguish if we had good gas-phase abundance measures of various isotopes... but maybe not if they aren't very stable....
 We discuss how this result may extend towards other metal yields that are not well sampled on our relatively short (500 Myr) simulation timescales in Section~\ref{sec:discussion:metal yields}.

@@

\subsection{Do These Results Apply to More Massive Galaxies?}
\label{sec:massive galaxies}
One important caveat about our work is that these results are derived from simulations of an isolated, low mass dwarf galaxy whose properties vary dramatically from a more massive galaxy like the Milky Way. It is unclear how much of these results apply to more massive galaxies given their deeper potential wells and higher star formation rates. Feedback driven galactic outflow properties do vary significantly with halo mass \citep[e.g.][]{MacLowFerrara1999,Muratov2017}, as more massive galaxies more easily retain and re-accrete gas. We expect the results presented in Figure~\ref{fig:species_fractions} to be the most susceptible to the particular star formation history and halo depth of a given galaxy. With gas from individual enrichment events more easily contained, we would expect the retention fractions to be more similar across metal species with increasing halo mass or decreasing star formation rate. If true, this difference in metal retention can be a key observable in verifying our result that the dynamical evolution of metals in the ISM is not uniform. If massive galaxies retain metals equally, we would expect dwarf galaxies to exhibit larger abundance ratios between AGB wind elements and SN enriched elements than stars in more massive galaxies, like the Milky Way. This difference would be greater at later times (higher metallicity) once AGB enrichment becomes significant. For example, at metallicities of [Fe/H] $\gtrsim$ 1, Ba is predominantly produced via s-process and ejected in AGB winds \citep{Travaglio1999,Travaglio2004}. At this metallicity, we would expect an elevated [Ba/$\alpha$] at fixed [$\alpha$/H] for low mass dwarfs as compared to the Milky Way. Indeed, this is found in the dSph's of the Milky Way \citep[see ][]{Tolstoy2009} and has been argued to either be the result of variations in s-process enrichment in metal poor environments and/or the result of enrichment timescale differences given the different star formation histories of dwarf galaxies and the Milky Way. In addtion, recent observations find elevationd [Ba/Fe] vs. [Fe/H] in local dwarf galaxies \citep{DugganKirbyAAS},
% AE TODO: Update duggand and kirby if / when the paper comes out
which could possibly be explained by this effect, but could also point towards significant Ba enrichment from NSMs. Given the lack of definitive explanation, clearly generalizing our results and detailing their observational consequences is a valuable area of future research.

@@

%
% moved to possible to-do list below,
%
\subsection{Implications for Exotic Enrichment Sources}
\label{sec:exotic enrichment}
We have shown that the dynamical evolution of metals depends upon their nucleosynthetic source, focusing here on AGB synthesized elements and supernova synthesized elements. However, these differences should also apply to exotic enrichment sources, such as hypernovae (HNe), neutron-star neutron-star mergers (NSM), and neutron star - black hole mergers. These sources have energies that differ significantly from typical supernova (10$^{51}$ erg), reaching $> 10^{52}$ for HNe \citep{Nomoto2004} and $\sim$10$^{49}$~erg for NSMs \citep{Goriely2011}. We would therefore expect different mixing behaviors for metals synthesized through these channels as compared to supernovae. Based on our results here, we would expect elements from HNe to be more well mixed in the ISM of all galaxies, but more readily ejected in dwarf galaxies, as compared to elements from SNe. Elements from NSMs, a significant source of r-process enrichment, would be less well-mixed in the ISM of all galaxies and more easily retained than SNe elements in dwarf galaxies. These differences could provide important signatures for distinguishing individual, exotic enrichment sources from observed stellar abundance patterns in our own Milky Way and in nearby dwarf galaxies. In particular, low mass dwarf galaxies with unusual r-process enrichment \citep[e.g.][]{Ji2015}, are valuable for constraining the source of these elements, their frequency, and typical yields. The differential metal evolution presented in our work both opens up an additional avenue by which elements from distinct nucleosynthetic sources may be distinguished in observations and challenges current assumptions used in interpreting these observations.

%\subsection{Extra things I could discuss:}
%\textit{Things I can / will additionally discuss}
%\begin{itemize}
%\item Observed N/O in stellar abundances exhibits large spreads at low metallicity. Argument has been that this is driven by timescale differences in AGB vs. massive star nucleosynthesis. However, our results would predict that N/O have a large spread due to mixing efficiency diffein rences between AGB winds and SNe.
%\end{itemize}

@@

\item Metal distributions exhibit positive skew, such that the median metal fraction can be anywhere from 0.1 to 1.0 dex above the mean metal fraction. Simple chemical evolution models, which generally follow the mean abundance, that are unable to account for complex metal mixing physics are likely biased towards higher enrichment values.

@@

We expect these results to be general for more massive galaxies and galaxy properties. However, it may be the case that metals couple more uniformly to galactic winds in more massive galaxies as wind driving occurs increasingly through entrainment of the ambient ISM rather than the ejection of individual enrichment events. This warrants further examination. But interpreting these results for low mass dwarf galaxies, we expect that: 1) s-process elements from AGB winds should exhibit larger spreads than $\alpha$ elements, particularly at lower metallicities, 2) these elements should be overabundant in dwarf galaxies at fixed age, as compared to massive galaxies like the Milky Way, and 3) that r-process enrichment from NS-NS mergers should behave similarly to s-process elements from AGB winds, as compared to $\alpha$ elements.

@@

\acknowledgments
Acknowledgments.

>>>>

\section{Introduction}
Understanding galactic chemical evolution for all metal species across galaxy mass scales remains one of the most challenging aspects of modeling galaxy evolution. One of the most pressing difficulties is a lack of understanding in exactly how metals propagate from their injection sites (from stellar winds or supernovae) and mix through the phases of the interstellar medium (ISM) into star forming gas.
%
% get a sentence with a better punch at the end
%

One-zone chemical evolution models assume homoegeneous mixing of metals from recent star formation into gas available for future star formation \citep[e.g.][]{Lanfranchi2006b,Kirby2011-III,Cote2017,Andrews2017}. While this incorrectly assumes that metal mixing in the ISM plays no role in delaying future enrichment in star formation, accounting for this process is challening. More complex models \citep[e.g.][]{SchonrichBinney2009,Pezzulli2016} employ multiple (usually radial) zones or follow multiple gas phases to attempt to account for these effects. How to properly account for multi-phase mixing, however, is still poorly understood. This is in large part because both these models and large-scale cosmological simulations lack the necessary fidelity to capture the detailed, multi-phase mixing process of metals in the ISM directly. Recent hydrodynamics simulations have employed parameteric models to account for unresolved sub-grid metal mixing \citep{PanScannapiecoScalo2013,Sarmento2017}, which plays an important role in determining the chemical properties of galaxies \citep[e.g.][]{Shen2010, Pilkington2012, Few2012, Brook2014, Escala2018}. Consequently, following metal mixing, if treated in these models, requires assumptions about how metals mix within the ISM and how metals couple to galactic outflows. When metals are tracked individually, as opposed to a global metallicity field, their injection, mixing, and outflow properties are often treated uniformly. However, these processes are critical components in understanding galactic chemical evolution. Detailed hydrodynamics simulations incorporating a realistic multi-phase ISM and detailed stellar feedback are required to understand metal mixing and metal outflows \citep[e.g.][]{FengKrumholz2014,Armillotta2018}. In addition, it remains to be seen to what degree, if at all, metals of different nucleosynthetic origins may couple differently to the ISM.

@@

It is becoming increasingly important to develop a concrete theoretical understanding of galactic chemical evolution as a result of multiple, recent observational campaigns to probe detailed stellar abundances in our Milky Way and the Local Group, such as SEGUE \citep{Yanny2009}, RAVE \citep{Kunder2017}, APOGEE \citep{Anders2014}, and GALAH \citep{Buder2018}. Stellar abundances are directly imprinted with the enrichment pattern of their star forming cloud, whose chemical properties are determined by the process of turbulent metal mixing in the ISM. The degree to which we can associate "chemically tagged" stars as co-eval depends directly upon our understanding of metal enrichment and metal mixing in the ISM. However, this understanding is critical for using these observations to further deduce properties of a galaxy's evolutionary history.

@@

\paragraph{Stellar Feedback and Stellar Yields} We track the feedback and yields of 15 metal species for each star individually. Stars between 8 M$_{\odot} < M_{*} < 25 M_{\odot}$ explode as core collapse supernovae at the end of their life, injecting their mass and 10$^{51}$~erg of thermal energy into a spherical region with radius of 5.4~pc, or 3 times the maximum resolution. Stars above this mass are assumed to direct collapse with no mass or energy injection. For all stars above 8~M$_{\odot}$ we follow their stellar winds assuming continuous mass loss over their lifetimes, their LW and FUV radiation as optically thin radiation which contributes to H$_2$ dissociation and photoelectric heating respectively, and their H~{\sc i} and He~{\sc i} ionizing radiation using an adaptive ray-tracing radiative transfer method \citep{WiseAbel2011}. We interpolate over the OSTAR2002 \citep{Lanz2003} grid to set the luminosities of each of these stars. Low mass stars, $M_{*} < 8~M_{\odot}$, do not produce feedback during their main sequence lifetimes, but end their lives injecting a short-lived, low velocity (10 km~s$^{-1}$) AGB wind \citep{Goldman2017}. Stars with $3 < M_{*} < 8$ are tracked after their death as possible Type Ia supernova progenitors, using a delay time distribution model to assign when (if at all) they will explode as a Type Ia, injecting 10$^{51}$~erg of thermal energy with yields. Stellar yields are computed using the NuGrid stellar yield database \citep{Ritter2017} for all stars with $M_{*} < 25~M_{\odot}$, \textbf{Slemer et. al. in prep} for the stellar winds of stars with $M_{*} > 25~M_{\odot}$, and \cite{Thielemann1986} for Type Ia supernovae.

@@

The only physics that separates the dynamical evolution of these elements in our simulations is the individual sources of enrichment: AGB winds (stars less than 8~M$_{\odot}$), stellar winds (stars above 8~M$_{\odot}$), core collapse supernovae, and Type Ia supernovae. These channels differ in: 1) how long after a given star formation event they occur, 2), by consequence, the typical ISM properties in which they occur, and 3) how much energy is associated with each event, which determines ejecta temperature and velocity. To understand where each of the elements in Figure~\ref{fig:species_fractions} originated, we show the mass fraction of metals produced through each channel in Figure~\ref{fig:species_sources}. Core collapse supernovae are responsible for over 90\% of the total metal enrichment in our galaxy, but this is clearly not true for all elements. In particular, a majority of N (74\%) and Ba (65\%) are ejected by AGB winds; both are retained at a higher rate than the rest of the metals. That this behavior exists for N and Ba in our simulations is dependent upon the metallicity of our galaxy and choice of stellar yield tables, but we can generalize this result to say that, for any assumed set of yields, low mass galaxies should more easily retain \textit{any} elements synthesized predominately in AGB winds, as compared to elements synthesized through supernovae.\footnote{One might expect that much of the N and Ba that is ejected by galactic winds is comprised mostly of the $\sim$30\% of each species produced through the stellar winds of more massive stars and supernovae. However, this cannot be proven as we lack Lagrangian information about individual gas elements.}
% AE: Actually if there are significant isotope variations between nucleosynthetic sources (i.e. N produced in AGB winds is a different isotope than the trace N produced in SNe) then that could be a very cool thing to distinguish if we had good gas-phase abundance measures of various isotopes... but maybe not if they aren't very stable....
We discuss how this result may extend towards other metal yields that are not well sampled on our relatively short (500 Myr) simulation timescales in Section~\ref{sec:discussion:metal yields}.

@@

\subsection{Do These Results Apply to More Massive Galaxies?}
\label{sec:massive galaxies}
One important caveat about our work is that these results are derived from simulations of an isolated, low mass dwarf galaxy whose properties vary dramatically from a more massive galaxy like the Milky Way. It is unclear how much of these results apply to more massive galaxies given their deeper potential wells and higher star formation rates. Feedback driven galactic outflow properties do vary significantly with halo mass \citep[e.g.][]{MacLowFerrara1999,Muratov2017}, as more massive galaxies more easily retain and re-accrete gas. We expect the results presented in Figure~\ref{fig:species_fractions} to be the most susceptible to the particular star formation history and halo depth of a given galaxy. With gas from individual enrichment events more easily contained, we would expect the retention fractions to be more similar across metal species with increasing halo mass or decreasing star formation rate. If true, this difference in metal retention can be a key observable in verifying our result that the dynamical evolution of metals in the ISM is not uniform. 
%If massive galaxies retain metals equally, we would expect dwarf galaxies to exhibit larger abundance ratios between AGB wind elements and SN enriched elements than stars in more massive galaxies, like the Milky Way.
This difference would be greater at later times, once AGB enrichment becomes significant. 
% Benoit suggested limiting / eliminating discussion relative to metallicity, since this is a little murky (age-metallicity varies significantly galaxy-to-galaxy). In addition, the elevated signals of Ba abundance ratios I mention have some amount of contention....
%For example, at metallicities of [Fe/H] $\gtrsim$ 1, Ba is predominantly produced via s-process and ejected in AGB winds \citep{Travaglio1999,Travaglio2004}. At this metallicity, we would expect an elevated [Ba/$\alpha$] at fixed [$\alpha$/H] for low mass dwarfs as compared to the Milky Way. Indeed, this is found in the dSph's of the Milky Way \citep[see ][]{Tolstoy2009} and has been argued to either be the result of variations in s-process enrichment in metal poor environments and/or the result of enrichment timescale differences given the different star formation histories of dwarf galaxies and the Milky Way. In addtion, recent observations find elevationd [Ba/Fe] vs. [Fe/H] in local dwarf galaxies \citep{DugganKirbyAAS}, which could possibly be explained by this effect, but could also point towards significant Ba enrichment from NSMs. Given the lack of definitive explanation, clearly generalizing our results and detailing their observational consequences is a valuable area of future research. 
If massive galaxies retain metals equally, we would expect dwarf galaxies to exhibit larger abundance ratios between AGB wind elements and SN enriched elements than stars in more massive galaxies, like the Milky Way.

@@

%
% moved to possible to-do list below,
%
\subsection{Implications for Exotic Enrichment Sources}
\label{sec:exotic enrichment}
We have shown that the dynamical evolution of metals depends upon their nucleosynthetic source, focusing here on AGB synthesized elements and supernova synthesized elements. However, these differences should also apply to exotic enrichment sources, such as hypernovae (HNe), neutron-star neutron-star mergers (NSM), and neutron star - black hole mergers. These sources can have energies that differ significantly from typical supernova (10$^{51}$ erg), reaching $> 10^{52}$ for HNe \citep{Nomoto2004}. In addition, the rarity of these events, approximately $10^{-5}$ per M$_{\odot}$ of star formation for NSMs \citep{Kim2015} in the Milky Way, could significantly influence the typical ISM environments within which they eject their metals. We would therefore expect different mixing behaviors for metals synthesized through these channels as compared to supernovae. Based on our results here, we would expect elements from HNe to be more well mixed in the ISM of all galaxies, but more readily ejected in dwarf galaxies, as compared to elements from SNe. Elements from NSMs, a significant source of r-process enrichment, would be less well-mixed in the ISM of all galaxies and more easily retained than SNe elements in dwarf galaxies. These differences could provide important signatures for distinguishing individual, exotic enrichment sources from observed stellar abundance patterns in our own Milky Way and in nearby dwarf galaxies. In particular, low mass dwarf galaxies with unusual r-process enrichment \citep[e.g.][]{Ji2016}, are valuable for constraining the source of these elements, their frequency, and typical yields. The differential metal evolution presented in our work both opens up an additional avenue by which elements from distinct nucleosynthetic sources may be distinguished in observations and challenges current assumptions used in interpreting these observations.

%\subsection{Extra things I could discuss:}
%\textit{Things to maybe}
%\begin{itemize}
%\item Observed N/O in stellar abundances exhibits large spreads at low metallicity. Argument has been that this is driven by timescale differences in AGB vs. massive star nucleosynthesis. However, our results would predict that N/O have a large spread due to mixing efficiency diffein rences between AGB winds and SNe.
%\end{itemize}

@@

\item Metal distributions exhibit positive skew, such that the mean metal fraction can be anywhere from 0.1 to 1.0 dex above the median metal fraction. Simple chemical evolution models, which generally follow the mean abundance, that are unable to account for complex metal mixing physics are likely biased towards higher enrichment values.

@@

We expect these results to be general for more massive galaxies and galaxy properties. However, it may be the case that metals couple more uniformly to galactic winds in more massive galaxies as wind driving occurs increasingly through entrainment of the ambient ISM rather than the ejection of individual enrichment events. This warrants further examination. But interpreting these results for low mass dwarf galaxies, we expect that: 1) s-process elements from AGB winds should exhibit larger spreads than $\alpha$ elements and 2) these elements should be overabundant in dwarf galaxies at fixed age, as compared to massive galaxies like the Milky Way.

@@

\acknowledgments
A.E. is funded by the NSF Graduate Research Fellowship DGE 16-44869. G.L.B. is funded by NSF AST-1312888, NASA NNX15AB20G, and NSF AST-1615955. M.-M.M.L. was partly funded by NASA  grant NNX14AP27G. B.C. acknowledges support from the ERC Consolidator Grant (Hungary) funding scheme (project RADIOSTAR, G.A. n. 724560) and from the National Science Foundation (USA) under grant No. PHY-1430152 (JINA Center for the Evolution of the Elements). \textbf{K.V.J....}. \textbf{B.O....}. We gratefully recognize computational resources provided by NSF XSEDE through grant number TGMCA99S024, the NASA High-End Computing Program through the NASA Advanced Supercomputing Division at Ames Research Center, Columbia University, and the Flatiron Institute. This work made significant use of many open source software packages, including \textsc{yt}, \textsc{Enzo}, \textsc{Grackle}, \textsc{Python}, \textsc{IPython}, \textsc{NumPy}, \textsc{SciPy}, \textsc{Matplotlib}, \textsc{HDF5}, \textsc{h5py}, \textsc{Astropy}, \textsc{Cloudy} and \textsc{deepdish}. These are products of collaborative effort by many independent developers from numerous institutions around the world. Their commitment to open science has helped make this work possible.
